---
title: "Understanding World Population Dynamics"
subtitle: "Assignment 1 - PSYC593"
quarto:
  outputs:
    figPath: ../results
format:
  html: default
date: today
author: Emily N. Line
---

<!-- These problems come from Imai, K. & Webb Williams, N. (2018). Quantitative Social Science: An introduction -->
<!-- in tidyverse. Princeton University Press. -->

<!-- The code giving partial solutions by Víctor H Cervantes -->



Understanding population dynamics is important for many areas of
social science.  We will calculate some basic demographic quantities
of births and deaths for the world's population from two time periods:
1950 to 1955 and 2005 to 2010.  We will analyze the following CSV data
files - `Kenya.csv`, `Sweden.csv`, and `World.csv`.  Each file
contains population data for Kenya, Sweden, and the world,
respectively.  The table below presents the names and descriptions of
the variables in each data set.

<!-- Note the syntax to create a table in markdown -->
<!-- Before the table there must be an empty line, and -->
<!-- the table begins with a line of dashes. -->
<!-- The column names are separated with spaces -->
<!-- and the first character must be a space -->
<!-- Then, segments of repeated dashes indicate the column separations (note -->
<!-- the space between groups of dashes before the character D of the column Description) -->

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `country`            Abbreviated country name
 
 `period`             Period during which data are collected
 
 `age`                Age group 
 
 `births`             Number of births in thousands (i.e., number of children 
                      born to women of the age group) 
 
 `deaths`             Number of deaths in thousands 
 
 `py.men`             Person-years for men in thousands 
 
 `py.women`           Person-years for women in thousands
--------------------------------------------------------------------------------

Source: United Nations, Department of Economic and Social Affairs, 
Population Division (2013). *World Population Prospects: The 2012 Revision, DVD Edition.*

```{r setup, message = FALSE}
# Load packages ----
library(tidyverse)
library(here)
library(flextable)


# Read data ----
# Using the function here() in order to specify location of file from 
# root directory
world_data <- readr::read_csv(here("data", "World.csv"))
kenya_data  <- readr::read_csv(here("data", "Kenya.csv"))
sweden_data <- readr::read_csv(here("data", "Sweden.csv"))

# Altering when Quarto translates something to scientific notation
options(scipen = 10)
```

<!-- The `read_csv` function from the package `readr` -->
<!-- is part of the `tidyverse`. It will read the data into a -->
<!-- special `data.frame` with more structure known as a `tibble` -->

The data are collected for a period of 5 years where *person-year* is
a measure of the time contribution of each person during the
period. For example, a person that lives through the entire 5 year
period contributes 5 person-years whereas someone who only lives
through the first half of the period contributes 2.5 person-years.
Before you begin this exercise, it would be a good idea to directly
inspect each data set.  In R, this can be done with the `View`
function, which takes as its argument the name of a `data.frame` to be
examined.  Alternatively, in RStudio, double-clicking a `data.frame`
in the `Environment` tab will enable you to view the data in a
spreadsheet-like view.

## Question 1

We begin by computing *crude birth rate* (CBR) for a given
  period.  The CBR is defined as:
  $$ 
    \text{CBR} 
     =  \frac{\text{number of births}}{\text{number of person-years lived}}
  $$
  <!-- It is also possible to include $\LaTeX$ syntax directly in the document. -->
  
  Compute the CBR for each period, separately for Kenya, Sweden, and
  the world.  Start by computing the total person-years, recorded as a
  new variable within each existing `data.frame` via the
  `$` operator, by summing the person-years for men and women.
  Then, store the results as a vector of length 2 (CBRs for two
  periods) for each region with appropriate labels.  You may wish to
  create your own function for the purpose of efficient programming.
  Briefly describe patterns you observe in the resulting CBRs.
  
  
  
### Answer 1


We calculate the total person years for each dataset by taking the sum of the
person years for men and the person years for women. This is stored in a new 
variable called `py_total`. 
```{r person-total}
# Create new variable py_total = total person years for each data set
world_data$py_total<-world_data$py.men+world_data$py.women
kenya_data$py_total<-kenya_data$py.men+kenya_data$py.women
sweden_data$py_total<-sweden_data$py.men+sweden_data$py.women
```


<!-- Whenever you find that you will repeat the same calculation -->
<!-- on different sets of data, **consider** creating a function -->
<!-- instead of repeating the code for each piece of data. -->

We create a function that calculated Crude Birth Rate (CBR), where the input
for the function is a dataframe that has a `births` variable and a `py_total` 
variable. The CBR is calculated by taking \frac{births}{py_total}. Separate CBRs
are calculated for each time period (in this case we have two separate periods).
```{r cbr-function}
# Function to compute the Crude Birth Rate (CBR)
# Takes a dataframe and computes the crude birth rate for two different periods
# of time.
# Assumes each dataset has births and py_total variables.
compute_cbr <- function (pop_data) {
  pop_data %>%
    group_by(period) %>%
    summarise(cbr = sum(births) / sum(py_total)) %>%
    pull()
}
```

<!-- The "pipe" `%>%` from package `magrittr` takes the output -->
<!-- from the left hand side and makes it the first argument of -->
<!-- the expression on the right hand side. -->

<!-- Starting in R 4.1.0, there is a base pipe operator `|>` -->
<!-- which has a similar behavior. -->

We calculate the CBR using the new function we just created, and we obtain two 
values for each dataset.
```{r cbr-computing, results = FALSE}
# Compute the CBR for each data set
(world_cbr <- compute_cbr(world_data))
(kenya_cbr <- compute_cbr(kenya_data))
(sweden_cbr <- compute_cbr(sweden_data))
```

**Answer** 

The CBR for Kenya in the time period 1950-1955 is `r round(kenya_cbr[1], 4)`. In the time
period 2005-2010, the CBR for Kenya is `r round(kenya_cbr[2], 4)`. The CBR for Sweden 
during 1950-1955 is `r round(sweden_cbr[1], 4)`. For the period of 2005-2010, Sweden's CBR 
is `r round(sweden_cbr[2], 4)`. The world's CBR for the period 1950-1955 is `r round(world_cbr[1], 4)`. The world's CBR for the period 2005-2010 is `r round(world_cbr[2], 4)`. 

  
## Question 2
  
The CBR is easy to understand but contains both men and women
   of all ages in the denominator. We next calculate the *total
     fertility rate* (TFR). Unlike the CBR, the TFR adjusts for age
   compositions in the female population.  To do this, we need to
   first calculate the *age specific fertility rate* (ASFR), which
   represents the fertility rate for women of the reproductive age
   range $[15, 50)$.  The ASFR for age range $[x, x+\delta)$, where
   $x$ is the starting age and $\delta$ is the width of the age range
   (measured in years), is defined as:
  $$
    \text{ASFR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of births to women of age $[x,\ x+\delta)$}}{\text{Number of person-years lived by women of age $[x,\ x+\delta)$}}
  $$
  Note that square brackets, $[$ and $]$, include the limit whereas
  parentheses, $($ and $)$, exclude it.  For example, $[20, 25)$
  represents the age range that is greater than or equal to 20 years
  old and less than 25 years old.  In typical demographic data, the
  age range $\delta$ is set to 5 years.  Compute the ASFR for Sweden
  and Kenya as well as the entire world for each of the two periods.
  Store the resulting ASFRs separately for each region.  What does the
  pattern of these ASFRs say about reproduction among women in Sweden
  and Kenya?
  

### Answer 2

The `compute_asfr` function removes age groups that are not applicable, and computes
the asfr for the remaining age groups and periods by taking \frac{births}{py.women}.
This creates a new variable in the dataset named `asfr`. 
```{r asfr-function}
# Function to compute Age specific fertility rate (ASFR)
# Removes age groups not applicable and creates new variable called asfr 
# by dividing births (during that period and for that age range) by the 
# women person-years 
compute_asfr <- function (pop_data) {
  pop_data %>%
    filter(age != "0-4" & age != "5-9" & age != "10-14" & age != "50-54" &
           age != "55-59" & age != "60-69" & age != "70-79" & age != "80+") %>% 
    mutate(asfr = births / py.women)
}
```

We then create the asfr variable using the `compute_asfr` function. We save it
to a new dataframe variable since we remove some age groups in the process. 
```{r region-asfr}
# Compute ASFR for each data set
world_asfr_df  <- compute_asfr(world_data)
kenya_asfr_df  <- compute_asfr(kenya_data)
sweden_asfr_df <- compute_asfr(sweden_data)
```
We now want to compare the ASFR's for Kenya and Sweden. By creating a scatterplot
to compare these values, we can better view how Sweden and Kenya might differ from 
each other. 
```{r asfr-plot}
# Compare ASFRs for Kenya and Sweden through a scatterplot
plot(kenya_asfr_df$asfr,
     sweden_asfr_df$asfr,
     xlab = "Kenya ASFR",
     ylab = "Sweden ASFR",
     xlim = c(0, 0.4),
     ylim = c(0, 0.4))

# title for plot
title("ASFR for Sweden and Kenya")

```
**Answer**

The scatterplot above shows the ASFR values for Sweden and Kenya. Looking at 
the pattern of data, we can see that Kenya consistenly has higher ASFR compared
to Sweden, indicating that the reproduction rates are much higher in Kenya than 
in Sweden. 


## Question 3

Using the ASFR, we can define the TFR as the average number of
  children women give birth to if they live through their entire
  reproductive age.
  $$
  \text{TFR} 
   =   \text{ASFR}_{[15,\ 20)} \times 5 + \text{ASFR}_{[20,\ 25)} \times 5 
  + \dots + \text{ASFR}_{[45,\ 50)} \times 5
  $$
  
  We multiply each age-specific fertility rate rate by 5 because the
  age range is 5 years.  Compute the TFR for Sweden and Kenya as well
  as the entire world for each of the two periods.  As in the previous
  question, continue to assume that women's reproductive age range is
  $[15, 50)$.  Store the resulting two TFRs for each country or the
  world as a vector of length two. In general, how has the number of
  women changed in the world from 1950 to 2000? What about the total
  number of births in the world?


### Answer 3

We create a function called `compute_tfr` that computes the TFR for each time
period, by multiplying the ASFR for that time period by 5 (because each age
range spans 5 years). This function gives a TFR value for each time period. 
```{r tfr-function}
# Function to compute the total fertility rate (TFR)
# Takes a dataset and computes the TFR for each time period by taking 5 * the total 
# asfr for that time period
compute_tfr <- function(pop_data) {
  pop_data %>%
    group_by(period) %>%
    summarise(tfr = 5 * sum(asfr)) %>%
    pull()
}
```

We then compute the TFR for each data set using the new `compute_tfr` function. 
```{r region-tfr, results = FALSE}
# Compute the TFR for each data set
(world_tfr  <- compute_tfr(world_asfr_df))
(kenya_tfr  <- compute_tfr(kenya_asfr_df))
(sweden_tfr <- compute_tfr(sweden_asfr_df))
```

We look at the total women in the world and the total births in the world
for each time period by taking the sum of each, grouped by time period. We can
then examine how these values have changed over time by taking the difference 
```{r world-tfr-totals, results = FALSE}
# Compute totals of women and births in the world by period
(
  totals_world <- world_data %>%
    group_by(period) %>%
    summarise(
      total_women = sum(py.women),
      total_births = sum(births)
    )
)

# Compare how much these totals have changed
# [total women/births from 2005-2010] / [total women/births from 1950-1955]
(changes_totals <- totals_world[2, -1] / totals_world[1, -1])
```
**Answer**

The total number of women in the world has increased from `r totals_world[[1, 2]]`
in 1950-1955 to `r totals_world[[2, 2]]` in 2005-2010. In other words, there are 
`r round(changes_totals$total_women[1], 2)` times the number of women in 2005-2010
compared to 1950-1955. 

The total number of births in the world (in the thousands) increased from `r totals_world[[1, 3]]`
in 1950-1955 to `r totals_world[[2, 3]]` in 2005-2010. There are `r round(changes_totals$total_births[1], 2)` times the number of births in 2005-2010
compared to 1950-1955. 

## Question 4

Next, we will examine another important demographic process:
  death.  Compute the *crude death rate* (CDR), which is a concept
  analogous to the CBR, for each period and separately for each
  region.  Store the resulting CDRs for each country and the world as
  a vector of length two. The CDR is defined as:
  $$ 
    \text{CDR} 
     =  \frac{\text{number of deaths}}{\text{number of person-years lived}}
  $$
  Briefly describe patterns you observe in the resulting CDRs.

We create a function that can calculate CDR per time period given a particular data set.
It does this by taking the total number of deaths within a time period divided 
by the total number of person-years within a time period. 
```{r cdr-function}
# Function to compute the Crude death rate (CDR)
# Groups by period to get CDR values for each time period.
# Calculates CDR by dividing all deaths in a period by 
# all person-years in a period
compute_cdr <- function (pop_data) {
  pop_data %>%
    group_by(period) %>%
    summarise(cbr = sum(deaths) / sum(py_total)) %>%
    pull()
}
```

We then use the new function `compute_cdr` to calculate CDR for each region. 
```{r region-cdr, results = FALSE}
# Compute the CDR for each data set
(world_cdr  <- compute_cdr(world_data))
(kenya_cdr  <- compute_cdr(kenya_data))
(sweden_cdr <- compute_cdr(sweden_data))
```

It's nice to view data in a table, so here we arrange the CDR data from each 
region into a table.
```{r cdr-table}
#| tbl-cap: CDR values per region and time period

# combining all regions into a single data set
cdr_data <- as.data.frame(rbind(world_cdr, kenya_cdr, sweden_cdr))

# Adding a column of region names
# Rounding all numbers to 5 decimal places
cdr_data2 <- cdr_data %>%
  add_column(Region = c("World", "Kenya", "Sweden"),
             .before = "V1") %>%
  mutate_if(is.numeric, round, digits = 5)

# creating a flextable object with the data
cdr_table <-
  regulartable(cdr_data2) %>%
  theme_zebra() %>%
  autofit() 

# Giving the columns informative titles
cdr_table <- set_header_labels(cdr_table,
                               values = list(V1 = "1950-1955",
                                             V2 = "2005-2010"))

# Printing the table (necessary for flextables in RMarkdown)
#cat(knitr::knit_print(cdr_table))
cdr_table
```

Viewing the table above, we can see that Kenya has a higher CDR than Sweden
and the "world average." This is true regardless of time period. We can also 
see that CDR values have all decreased in the more recent time period
compared to the older time period. Sweden has better CDRs than Kenya and the 
world average in 1950-1955, but has a slightly higher CDR than the world average
in 2005-2010. 

## Question 5

One puzzling finding from the previous question is that the CDR
  for Kenya during the period of 2005-2010 is about the same level as
  that for Sweden.  We would expect people in developed countries like
  Sweden to have a lower death rate than those in developing countries
  like Kenya.  While it is simple and easy to understand, the CDR does
  not take into account the age composition of a population.  We
  therefore compute the *age specific death rate* (ASDR).  The
  ASDR for age range $[x, x+\delta)$ is defined as:
  $$
    \text{ASDR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of deaths for people of age $[x,\ x+\delta)$}}
               {\text{number of person-years of people of age $[x,\ x+\delta)$}}
  $$
  Calculate the ASDR for each age group, separately for Kenya and
  Sweden, during the period of 2005-2010.  Briefly describe the
  pattern you observe.
 

We create a function called `compute_asdr` that first filters the data set so 
that only the time period 2005-2010 remains. Then it creates a new variable
called `asdr` that is calculated by taking the deaths divided by the total
number of person-years for each age range. 
```{r asdr-function}
# Function to compute Age specific death rate (ASDR)
# First removes any time periods that are not 2005-2010
# Then takes a data set and computes asdr by dividing 
# number of deaths by person-year total 
compute_asdr <- function (pop_data) {
  pop_data %>%
    filter(period == "2005-2010") %>% 
    mutate(asdr = deaths / py_total)
}
```

We then use the new `compute_asdr` function to calculate ASDR for each region. 
```{r region-asdr}
# Compute ASDR for each data set
world_asdr  <- compute_asdr(world_data)
kenya_asdr  <- compute_asdr(kenya_data)
sweden_asdr <- compute_asdr(sweden_data)
```

It's easier to compare the ASDR values between Sweden and Kenya when we use a 
scatterplot. This code creates a scatterplot comparing Sweden's ASDR and Kenya's
ASDR. 
```{r asdr-plot}
# Compare ASDRs for Kenya and Sweden through a scatterplot
plot(kenya_asdr$asdr,
     sweden_asdr$asdr,
     xlab = "Kenya ASDR",
     ylab = "Sweden ASDR",
     xlim = c(0, 0.014),
     ylim = c(0, 0.014))

# title for plot
title("ASDR for Sweden and Kenya")
```
 
The scatterplot above shows that Kenya has much larger ASDR values than Sweden. 
 
## Question 6

One way to understand the difference in the CDR between Kenya
  and Sweden is to compute the counterfactual CDR for Kenya using
  Sweden's population distribution (or vice versa).  This can be done
  by applying the following alternative formula for the CDR.
  $$
    \text{CDR}
    \ = \ \text{ASDR}_{[0, 5)} \times P_{[0,5)} + \text{ASDR}_{[5, 10)}
    \times P_{[5, 10)} + \cdots 
  $$
  where $P_{[x, x+\delta)}$ is the proportion of the population in the
  age range $[x, x+\delta)$.  We compute this as the ratio of
  person-years in that age range relative to the total person-years
  across all age ranges.  To conduct this counterfactual analysis, we
  use $\text{ASDR}_{[x,x+\delta)}$ from Kenya and $P_{[x,x+\delta)}$
  from Sweden during the period of 2005--2010.  That is, first
  calculate the age-specific population proportions for Sweden and
  then use them to compute the counterfactual CDR for Kenya. How does
  this counterfactual CDR compare with the original CDR of Kenya?
  Briefly interpret the result.


We create a function called `compute_pop_prop` that will compute the population 
proportion for the time period 2005-2010 by taking the person-years total 
for each age range divided by the sum of all person-years for that period. 
```{r pop-prop-function}
# Function to compute population proportion by period
# First filter out time periods other than 2005-2010. 
# Calculates population proportion by dividing 
# person-years total for that age group by the sum of
# person-years for that period. 
compute_pop_prop <- function (pop_data) {
  pop_data %>%
    filter(period == "2005-2010") %>% 
    mutate(pop_prop = py_total / sum(py_total)) %>%
    ungroup()
}
```

We use the new `compute_pop_prop` function to calculate the population proportion
per time period for each region. 
```{r region-pop-prop}
# Compute population proportion for each data set
world_pop <- compute_pop_prop(world_data)
kenya_pop <- compute_pop_prop(kenya_data)
sweden_pop <- compute_pop_prop(sweden_data)
```

Lastly, we calculate the counterfactual CDR for Kenya by multiplying Kenya's
ASDR by Sweden's population proportion, for each age group. Then we take the
sum of those values within each time period to obtain a period-specific 
counterfactual CDR for Kenya with respect to Sweden. 
```{r kenya-counterfactual, results = FALSE}
# Compute Kenyas CDR Kenya had Sweden's population distribution
kenya_counter_cdr <-
  mutate(kenya_asdr,
         temp_cdr  = asdr * sweden_pop$pop_prop) %>%
  summarise(cdr_re_sweden = sum(temp_cdr))
```

The counterfactual CDR value for Kenya in 2005-2010 is `r round(kenya_counter_cdr, 4)`,
which is higher than the original Kenya CDR value of `r round(kenya_cdr[2], 4)`. 
This suggests that if Kenya had the age distribution of Sweden in 2005-2010, the
CDR would be even higher. 